### 2025-02-18 paper/relatedwork.tex
Cut.
Some more related work.

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly? (ii) if the
language or the framework do not perform as expected, what are the
options to solve this?  Unfortunately, for many cases the answer to
both questions is unsatisfying.  Most of the languages/frameworks do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.  Furthermore, these systems
often rely on sophisticated 
combinations of sub-systems that need to be fine-tuned to the executing hardware.

@@

>>>>

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly? (ii) if the
language or the framework do not perform as expected, what are the
options to solve this?  Unfortunately, for many cases the answer to
both questions is unsatisfying.  Most of the languages/frameworks do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.

@@

Verified high-performance computing is an area of active research, and
one line of research is based on the idea of separating the high level
\emph{specification} of a problem from the \emph{schedule} that
describe how it is to be executed. Verifying that such schedules are
semantics-preserving is the topic of several recent
publications~\cite{10.1145/3527328,10.1145/3498717}. Our approach
differs by not using a scheduling language, and instead focuses on
verifying the specification itself, and also by having a particularly
expressive specification language.
